Semi-Supervised Factored Logistic Regression for High-Dimensional Neuroimaging Data

### 2015-06-05 paper_nips/paper.tex
included OGs comments

Both these neurobiological properties can be conjointly exploited in an
approach that is
\textit{mixed} (i.e., using rest and task data),
\textit{factored} (i.e., performing network decomposition), and
\textit{multi-task} (i.e., capitalize on neural representations shared across
mental operations).
%
The integration of brain-network discovery into
supervised classification can yield a semi-supervised learning
framework. The most relevant
neurobiological structure should hence be identified
for the prediction problem at hand.
%
Autoencoders suggest themselves because they can emulate
variants of most unsupervised learning algorithms,
including PCA, SPCA, and ICA \cite{hinton06}.
\begin{wrapfigure}{r}{0.40\textwidth}
  \centering
  %\captionsetup{labelformat=nonumber}
    \includegraphics[width=0.40\textwidth]{figures/figure1.png}
  \vspace{-0.7cm}
  \caption {\textbf{Model architecture.}
  Linear autoencoders find an optimized compression of 79,941 brain voxels
  into $n$ unknown activity patterns by improving reconstruction from them.
  The decomposition matrix equates with the bottleneck of
  a factored logistic regression.
  Supervised multi-class learning on task data ($X_{task}$)
  can thus be guided by
  unsupervised decomposition of rest data ($X_{rest}$).
  }
\end{wrapfigure}
%
Autoencoders (AE)
are layered learning models
that condense the input data to
local and global 
representations
via reconstruction under compression prior.
%
% XXX Michael: In fact, they behave like a non-centered PCA, 
% XXX which amounts to just doing an SVD
% XXX which btw is NOT a convex optimization problem, although easy to 
% XXX calculate and with a global optimum.
% XXX Autoencoders are also not one-layered by default. Let's make sure that
% XXX we don't trip up by getting the technicalities wrong and getting 
% XXX reviewers annoyed at us because of it.
% DanilO: cool thanks. That is the stuff, we need.
They behave like a (truncated) PCA
in case of one linear hidden layer and a squared error loss
\cite{baldi1989neural}.
% This architecture yields an optimization objective
% with unique global minimum.
Autoencoders behave like a SPCA if shrinkage terms are added to the
model weights in the optimization objective.
Moreover, they have the characteristics of an ICA in case of tied weights
and adding a nonlinear convex function
at the first layer \cite{le2011ica}.
% XXX What convex transformation? Why is that relevant? Convexity is only
% XXX relevant for the loss
% Danilo: it's about adding a neural-network-like nonconvex function for
% first-layer activation
These authors further demonstrated that ICA, sparse autoencoders, and 
sparse coding are mathematically equivalent
under mild conditions.
%
Thus, autoencoders may flexibly project the neuroimaging data
onto the main directions of variation.

In the present investigation,
a linear autoencoder will be fit to
(unlabeled) rest data and
integrated as a rank-reducing bottleneck
into a multinomial
logistic regression fit to
(labeled) task data.
We can then solve the unsupervised data representation and the
supervised classification, previously studied in isolation,
in an identical statistical learning problem.
%
From the perspective of dictionary learning, the first layer represents
% XXX Michael: This is the crucial difference: The first layer is an analysis
% XXX operation: it extracts information in some way. It may be extracting
% XXX activations for basis functions, in which case it would be the pseudo-
% XXX inverse of those basis functions. But unless they are orthogonal, these
% XXX weights definitely don't represent basis functions themselves. It is 
% XXX the second layer of the autoencoder that represents basis functions. If
% XXX weights are tied, they are one and the same thing, with interesting
% XXX implications. But in general, only the second layer represents a basis.
projectors to the discovered set of basis functions
which are linearly combined 
by the second layer to perform predictions \cite{olshausen96}.
% XXX Michael: ^ This needs to be changed around: This first layer obtains
% XXX activations, the second layer weights are basis functions which are
% XXX multiplied to these activations
%
Neurobiologically, this allows 
delineating a low-dimensional manifold
of brain network patterns and then 
distinguishing mental tasks
by their most discriminative linear combinations.
%
Theoretically, a reduction in model variance should be achieved by
resting-state autoencoders that
privilege the most neurobiologically
valid models in the hypothesis set.
%
Practically, neuroimaging research frequently suffers from
data scarcity. This limits the set of representations that can be
extracted from GLM analyses based on few participants.
%
We therefore contribute a computational framework that
1) analyzes many problems simultaneously
(thus finds shared representations by ``multi-task learning'')
and
2) exploits unlabeled data,
as they span a space of meaningful configurations.

\section{Methods}
%
\paragraph{Data.}
As the currently biggest openly-accessible reference dataset,
we chose resources from the Human Connectome Project (HCP)
\cite{barch2013}.
Neuroimaging task data with labels of ongoing cognitive processes
were drawn from 500 unrelated,
healthy HCP participants (cf. Appendix for details on datasets).
18 HCP tasks 
were selected that are known to elicit reliable neural activity
across participants (Table \ref{table_tasks}).
In sum, the HCP task data incorporated 8650 first-level activity maps
from 18 diverse paradigms administered to 498 participants (2 removed
due to incomplete data).
All maps were resampled to a common $60\times72\times60$ space of
3mm isotropic voxels and gray-matter masked (at least 10\% tissue
probability).
The supervised analyses were based on labeled HCP task maps with
79,941 voxels of interest representing z-values in gray matter.

\begin{table}[h]
  \resizebox{0.98\textwidth}{!}{%
  \begin{tabular}{l|l|l}
    \hline
  {\bf Cognitive Task} & {\bf Stimuli}                         & {\bf Instruction for participants}                                                \\ \hline
  1 Reward             & \multirow{2}{*}{Card game}            & \multirow{2}{*}{Guess the number of a mystery card for gain/loss of money}        \\ \cline{1-1}
  2 Punish             &                                       &                                                                                   \\ \hline
  3 Shapes             & Shape pictures                        & Decide which of two shapes matches another shape geometrically                    \\ \hline
  4 Faces              & Face pictures                         & Decide which of two faces matches another face emotionally                        \\ \hline
  5 Random             & \multirow{2}{*}{Videos with objects}  & \multirow{2}{*}{Decide whether the circles, squares, triangles act intentionally} \\ \cline{1-1}
  6 Theory of mind     &                                       &                                                                                   \\ \hline
  7 Mathematics        & Spoken numbers                        & Complete addition and subtraction problems                                        \\ \hline
  8 Language           & Auditory stories                      & Choose answer about the topic of the story                                        \\ \hline
  9 Tongue movement    & \multirow{3}{*}{Visual cues}          & Move tongue                                                                       \\ \cline{1-1} \cline{3-3} 
  10 Food movement     &                                       & Squeezing of the left or right toe                                                \\ \cline{1-1} \cline{3-3} 
  11 Hand movement     &                                       & Tapping of the left or right finger                                               \\ \hline
  12 Matching          & \multirow{2}{*}{Shapes with textures} & Decide whether two objects match in shape or texture                             \\ \cline{1-1} \cline{3-3} 
  13 Relations         &                                       & Decide whether object pairs differ both along either shape or texture             \\ \hline
  14 View Bodies       & Pictures                              & Passive watching                                                                   \\ \hline
  15 View Faces        & Pictures                              & Passive watching                                                                   \\ \hline
  16 View Places       & Pictures                              & Passive watching                                                                   \\ \hline
  17 View Tools        & Pictures                              & Passive watching                                                                   \\ \hline
  18 Two-Back          & Various pictures                      & Indicate whether current stimulus is the same as two items earlier                \\ \hline
  \end{tabular}
}
\vspace{-0.2cm}
\caption{\textbf{Description of psychological tasks to predict.}}
\label{table_tasks}
\end{table}

@@

We were further interested in the utility of the
optimized low-rank projection
in one task dataset for dimensionality reduction in another task dataset.
To this end, the HCP-derived network decompositions were used as preliminary
step in the classification problem of another large sample.
The ARCHI dataset \cite{pinel07} provides activity maps from
diverse experimental tasks, including auditory and visual perception, motor action,
reading, language comprehension and mental calculation.
Analogous to HCP data, the second task dataset incorporated 1404
labeled, grey-matter masked, and z-scored activity maps
from 18 diverse tasks acquired in 78 participants.

The labeled and unlabeled data were fed into a linear statistical model
composed of an autoencoder and dimensionality-reducing logistic regression.

\paragraph{Linear autoencoder.}
The affine autoencoder takes the input 
$\mathbf{x}$ and projects it into a coordinate system of
latent representations $\mathbf{z}$
and reconstructs it back to $\mathbf{x'}$ by
\begin{eqnarray}
    \mathbf{z} = \mathbf{W_0} \mathbf{x} + \mathbf{b_0} \hspace{2cm}
    \mathbf{x'} = \mathbf{W_1} \mathbf{z} + \mathbf{b_1},
  \label{eq:autoenc}
\end{eqnarray}

@@

%with \(f_{AE}(\mathbf x) = %
%\mathbf{W_1}(\mathbf{W_0}\mathbf{x} + \mathbf{b_0}) + \mathbf{b_1}\)
% the autoencoder.
% XXX Michael: I don't know to what extent the following makes sense or is
% XXX correct, so I am commenting it
% XXX Danilo: I found this in Bengio's papers somewhere
This reconstruction error criterion equates with
maximizing a lower bound on the mutual information between
input and the learned representation.
% XXX Give a ref ?
Here we choose $\mathbf{W_0}$ and $\mathbf{W_1}$ to be tied, i.e.
$\mathbf{W_0} = \mathbf{W_1^T}$. In particular, this means that the weights
learned are forced to take a two-fold function: That of signal 
\textit{analysis} and that of signal \textit{synthesis}.
The first layer \textit{analyzes}
the data to obtain the cleanest latent representation,
while the second
layer represents building blocks from which to \textit{synthesize} the data
using the latent activations.
Tying these processes together makes the analysis
layer interpretable and pulls all non-zero singular values towards 1. 
Nonlinearities were not applied to the
activations in the first layer.

\paragraph{Factored logistic regression.}
Lossy compression by a low-dimensional bottleneck
is also imposed by the first layer of the factored
multinomial logistic regression.
It gives the probability of an input $\mathbf{x}$ to belong
to a class $i \in \{1, \dots, l\}$
\begin{equation}
  \begin{split}
    P(Y=i|\mathbf{x; V_0,V_1,c_0, c_1}) &= \softmax_i(f_\mathcal{LR}(\mathbf x)),
  \end{split}
  \label{eq:lr}
\end{equation}
where \(f_\mathcal{LR}(\mathbf x) = \mathbf{V_1 (V_0 x + c_0) + c_1}\) computes 
multinomial logits and \(\softmax_i(x) = \exp(x_i)/\sum_j\exp(x_j)\).
The matrix $\mathbf{V_0 \in \mathbb{R}^{dxn}}$
transforms the input $\mathbf{x \in \mathbb{R}^{d}}$
into $n$ latent components
and the matrix $\mathbf{V_1 \in \mathbb{R}^{nxl}}$
projects the latent components
onto hyperplanes that reflect $l$ label probabilities.
$\mathbf{c_0}$ and $\mathbf{c_1}$ are
bias vectors.
The loss function is given by
\begin{equation}
    \mathbb E\left[{\mathcal{L_{LR}}}(\mathbf{x, y})\right] \approx %
\frac{1}{N_{X_{task}}} \sum_{k=0}^{N_{X_{task}}} \log(P(Y=y^{(k)}|\mathbf{x^{(k)}; V_0, V_1, c_0, c_1}).
\label{eq:lr_loss}
\end{equation}

\paragraph{Layer combination.}
Importantly, the optimization problem of the linear autoencoder
and the factored logistic regression
are linked on two levels. First, their transformation matrices mapping from
input to the latent space are tied, i.e.
\begin{eqnarray}
  \mathbf{V_0} = \mathbf{W_0}.
\end{eqnarray}
We thus search for a compression of the 79,941 voxel values into $n$ latent
components that represent a latent code optimized for both
rest and task activity data.
Second, the objectives of the autoencoder and the factored
logistic regression are interpolated in the common loss function
% XXX Michael: In resolving conflict I opted for my change. Feel free to revert if that was intended.
\begin{equation}
{\mathcal{L}}(\theta, \lambda) = \lambda {\mathcal{L_{LR}}}
+ (1-\lambda)\frac{1}{N_{X_{rest}}} {\mathcal{L_{AE}}} + \Omega(\theta).
  \label{eq:loss_equ}
\end{equation}

@@

\paragraph{Optimization.}
The common objective was optimized
by gradient descent in the SSFLogReg parameters.
%^
The required gradients were obtained by using the chain rule to
backpropagate error derivatives.
We chose the \textit{rmsprop} solver \cite{rmsprop},
a refinement of stochastic gradient descent.
\textit{Rmsprop} dictates an adaptive learning rate
for each model parameter by
scaled gradients from a running average.
The batch size was set to $100$
(much expected redundancy in $X_{rest}$ and $X_{task}$),
matrix parameters were initalized by Gaussian random values with
standard deviation $0.004$ (i.e., gain), and
bias parameters were initalized to $0$.

The normalization factor and the update rule for $\theta$
are given by
% XXX Michael: It is crucial, but nowhere mentioned, that the optimization is
% XXX done in an alternating manner. The whole structure relies on this, by
% XXX construction it is vulnerable to optimizing too much towards labels or
% XXX rest at a time. I also identify this as one of the major weak points
% XXX of the method, so it needs to be stated in a way that is not easily
% XXX attackable.
% XXX Maybe we could also talk less about RMSprop.
%
% XXX Danilo: I think we are not alternating, Xrest and Xtask go in
% simultaneously at every bach in every epoch
\begin{eqnarray}
  \begin{split}
    \mathbf{v^{(t+1)}} &= \rho \mathbf{v^{(t)}} + (1 - \rho)\left(\nabla_{\theta} f(x^{(t)}, y^{(t)}, \theta^{(t)})\right)^2
% XXX: I have changed the first equation, please check
%v^{(t+1)} = \rho v^{(t+1)} + (1 - \rho)˜|\nabla f
\\
\theta^{(t+1)} &= \theta^{(t)} + \alpha \frac{\nabla_{\theta} f(x^{(t)}, y^{(t)}, \theta^{(t)})}{\sqrt{\mathbf{v^{(t+1)}} + \epsilon}},
  \end{split}
\end{eqnarray}
% XXX Michael: Why, in the formula, do we have two different notations for
% XXX partial derivatives? \partial and \nabla
% XXX Michael: Probably remove the explanations following and just state values
% XXX: Bertrand: what is f ?? Danilo: replaced f by more detailed notation
where $f$ is the loss function computed on a minibatch sample at timestep $t$,
$\alpha$ is the learning rate ($0.00001$),
$\epsilon$ a global damping factor ($10^{-6}$)
and
$\rho$ the decay rate ($0.9$ to deemphasize the magnitude of the gradient).
%
Note that we have also experimented with other solvers
(stochastic gradient descent, adadelta, and adagrad) but found that
\textit{rmsprop} converged faster and with
similar or higher generalization performance.

\paragraph{Hints.}
In fact, the constraint by a rest-data autoencoder qualifies as a
\textit{hint}
rather than regularization in a strict sense \cite{abu1994hints}.
Its purpose is to introduce
% XXX Michael: I use f earlier up, is this f here necessary?
prior information on
\textit{known} properties of the \textit{unknown} target function $f$.
Rather than only relying on input-output pairs in the learning process,
we thus narrow our hypothesis set to the biologically most plausible solutions.
That is, we reduce the search space in a way that
is compatible with the expected representation of BOLD activity signals.

\paragraph{Implementation.}
The analyses were performed in Python.
We used \textit{nilearn} to handle
the large quantities of neuroimaging data 
\cite{abrah14}
and
\textit{Theano} for automatic, numerically stable
differentiation of symbolic computation graphs
\cite{bastien2012theano, bergstra2010theano}.
All Python scripts that generated the results are
accessible online for reproducibility and reuse
\url{http://github.com/anonymous/anonymous}.
% url{http://github.com/banilo/nips2015}.

@@

SPCA separated the BOLD signals into

@@

We made several noteworthy observations (Table \ref{table_two}).
%
First, the most supervised estimator ($\lambda=1$) achieved in no
instance the best accuracy, precision, recall or f1 scores on HCP data.
Classification by SSFLogReg is therefore facilitated by
imposing structure from the unlabeled rest data.
%
Second, the higher the number of latent components $n$,
the higher the out-of-study performance with small values of $\lambda$.
This suggests that the presence of more rest-data-inspired hidden components
results in more effective feature representations in unrelated task data.
%
Third, for $n=20$ and $100$ (but not $5$) the purely rest-data-trained
decomposition matrix ($\lambda=0$) resulted in
noninferior out-of-study performance
of 77,0\% and 79,2\%,
respectively (Table \ref{table_two}).
This confirms that guiding model learning by task-unrelated structure
extracts features of general relevance
beyond the supervised problem at hand.

@@

We were further interested in the utility of the
optimized low-rank projection
in one task dataset for dimensionality reduction in another task dataset.
To this end, the HCP-derived network decompositions were used as preliminary
step in the classification problem of another large sample.
The ARCHI dataset \cite{pinel07} provides activity maps from
diverse experimental tasks, including auditory and visual perception, motor action,
reading, language comprehension and mental calculation.
81 right-handed healthy participants
(3 not included in present analyses due to incomplete data)
without psychiatric or
neurological history participated in four fMRI sessions acquired under
different experimental paradigms.
The functional maps were warped into
the MNI space and resampled to isotropic 3mm resolution.
Whole-brain EPI data were acquired with the same Siemens Trio with a 32
channel head coil (TR=2400ms, TE=30ms, flip angle=60°, in-plane
FOV=$192\textrm{mm}\times192\textrm{mm}$, 40 slices, 3.0mm isotropic voxels).
Standard preprocessing was performed with Nipype \cite{gorgo11}, including
slice timing, motion correction, alignment, and spatial normalization.
Activity maps were spatially smoothed by
a Gaussian kernel of 5mm (FWHM).
Analogous to HCP data, the second task dataset incorporated 1404
labeled, grey-matter masked, and z-scored activity maps
from 18 diverse tasks acquired in 78 participants.

>>>>

Both these neurobiological properties can be conjointly exploited in an
approach that is
\textit{mixed} (i.e., using rest and task data),
\textit{factored} (i.e., performing network decomposition), and
\textit{multi-task} (i.e., capitalize on neural representations shared across
mental operations).
%
The integration of brain-network discovery into
supervised classification can yield a semi-supervised learning
framework. The most relevant
neurobiological structure should hence be identified
for the prediction problem at hand.
%
Autoencoders suggest themselves because they can emulate
variants of most unsupervised learning algorithms,
including PCA, SPCA, and ICA \cite{hinton06}.
\begin{wrapfigure}{r}{0.40\textwidth}
  \centering
  %\captionsetup{labelformat=nonumber}
    \includegraphics[width=0.40\textwidth]{figures/figure1.png}
  \vspace{-0.7cm}
  \caption {\textbf{Model architecture}
  Linear autoencoders find an optimized compression of 79,941 brain voxels
  into $n$ unknown activity patterns by improving reconstruction from them.
  The decomposition matrix equates with the bottleneck of
  a factored logistic regression.
  Supervised multi-class learning on task data ($X_{task}$)
  can thus be guided by
  unsupervised decomposition of rest data ($X_{rest}$).
  }
\end{wrapfigure}
%
Autoencoders (AE)
are layered learning models
that condense the input data to
local and global 
representations
via reconstruction under compression prior.
%
% XXX Michael: In fact, they behave like a non-centered PCA, 
% XXX which amounts to just doing an SVD
% XXX which btw is NOT a convex optimization problem, although easy to 
% XXX calculate and with a global optimum.
% XXX Autoencoders are also not one-layered by default. Let's make sure that
% XXX we don't trip up by getting the technicalities wrong and getting 
% XXX reviewers annoyed at us because of it.
% DanilO: cool thanks. That is the stuff, we need.
They behave like a (truncated) PCA
in case of one linear hidden layer and a squared error loss
\cite{baldi1989neural}.
% This architecture yields an optimization objective
% with unique global minimum.
Autoencoders behave like a SPCA if shrinkage terms are added to the
model weights in the optimization objective.
Moreover, they have the characteristics of an ICA in case of tied weights
and adding a nonlinear convex function
at the first layer \cite{le2011ica}.
% XXX What convex transformation? Why is that relevant? Convexity is only
% XXX relevant for the loss
% Danilo: it's about adding a neural-network-like nonconvex function for
% first-layer activation
These authors further demonstrated that ICA, sparse autoencoders, and 
sparse coding are mathematically equivalent
under mild conditions.
%
Thus, autoencoders may flexibly project the neuroimaging data
onto the main directions of variation.

In the present investigation,
a linear autoencoder will be fit to
(unlabeled) rest data and
integrated as a rank-reducing bottleneck
into a multinomial
logistic regression fit to
(labeled) task data.
We can then solve the compound statistical problem of
unsupervised data representation and
supervised classification, previously studied in isolation.
%
From the perspective of dictionary learning, the first layer represents
% XXX Michael: This is the crucial difference: The first layer is an analysis
% XXX operation: it extracts information in some way. It may be extracting
% XXX activations for basis functions, in which case it would be the pseudo-
% XXX inverse of those basis functions. But unless they are orthogonal, these
% XXX weights definitely don't represent basis functions themselves. It is 
% XXX the second layer of the autoencoder that represents basis functions. If
% XXX weights are tied, they are one and the same thing, with interesting
% XXX implications. But in general, only the second layer represents a basis.
projectors to the discovered set of basis functions
which are linearly combined 
by the second layer to perform predictions \cite{olshausen96}.
% XXX Michael: ^ This needs to be changed around: This first layer obtains
% XXX activations, the second layer weights are basis functions which are
% XXX multiplied to these activations
%
Neurobiologically, this allows 
delineating a low-dimensional manifold
of brain network patterns and then 
distinguishing mental tasks
by their most discriminative linear combinations.
%
Theoretically, a reduction in model variance should be achieved by
resting-state autoencoders that
privilege the most neurobiologically
valid models in the hypothesis set.
%
Practically, neuroimaging research frequently suffers from
data scarcity. This limits the set of representations that can be
extracted from GLM analyses based on few participants.
%
We therefore contribute a computational framework that
1) analyzes many problems simultaneously
(thus finds shared representations by ``multi-task learning'')
and
2) exploits unlabeled data,
as they span a space of meaningful configurations.

\section{Methods}
%
\paragraph{Data.}
As the currently biggest openly-accessible reference dataset,
we chose resources from the Human Connectome Project (HCP)
\cite{barch2013}.
Neuroimaging task data with labels of ongoing cognitive processes
were drawn from 500 unrelated,
healthy HCP participants (cf. Appendix for details on datasets).
18 HCP tasks 
were selected that are known to elicit reliable neural activity
across participants (Table \ref{table_tasks}).
In sum, the HCP task data incorporated 8650 first-level activity maps
from 18 diverse paradigms administered to 498 participants (2 removed
due to incomplete data).
All maps were resampled to a common $60\times72\times60$ space of
3mm isotropic voxels and gray-matter masked (at least 10\% tissue
probability).
The supervised analyses were thus based on labeled HCP task maps with
79,941 voxels of interest representing z-values in gray matter.

\begin{table}[h]
  \resizebox{0.98\textwidth}{!}{%
  \begin{tabular}{l|l|l}
    \hline
  {\bf Cognitive Task} & {\bf Stimuli}                         & {\bf Instruction for participants}                                                \\ \hline
  1 Reward             & \multirow{2}{*}{Card game}            & \multirow{2}{*}{Guess the number of a mystery card for gain/loss of money}        \\ \cline{1-1}
  2 Punish             &                                       &                                                                                   \\ \hline
  3 Shapes             & Shape pictures                        & Decide which of two shapes matches another shape geometrically                    \\ \hline
  4 Faces              & Face pictures                         & Decide which of two faces matches another face emotionally                        \\ \hline
  5 Random             & \multirow{2}{*}{Videos with objects}  & \multirow{2}{*}{Decide whether the objects act randomly or intentionally} \\ \cline{1-1}
  6 Theory of mind     &                                       &                                                                                   \\ \hline
  7 Mathematics        & Spoken numbers                        & Complete addition and subtraction problems                                        \\ \hline
  8 Language           & Auditory stories                      & Choose answer about the topic of the story                                        \\ \hline
  9 Tongue movement    & \multirow{3}{*}{Visual cues}          & Move tongue                                                                       \\ \cline{1-1} \cline{3-3} 
  10 Food movement     &                                       & Squeezing of the left or right toe                                                \\ \cline{1-1} \cline{3-3} 
  11 Hand movement     &                                       & Tapping of the left or right finger                                               \\ \hline
  12 Matching          & \multirow{2}{*}{Shapes with textures} & Decide whether two objects match in shape or texture                             \\ \cline{1-1} \cline{3-3} 
  13 Relations         &                                       & Decide whether object pairs differ both along either shape or texture             \\ \hline
  14 View Bodies       & Pictures                              & Passive watching                                                                   \\ \hline
  15 View Faces        & Pictures                              & Passive watching                                                                   \\ \hline
  16 View Places       & Pictures                              & Passive watching                                                                   \\ \hline
  17 View Tools        & Pictures                              & Passive watching                                                                   \\ \hline
  18 Two-Back          & Various pictures                      & Indicate whether current stimulus is the same as two items earlier                \\ \hline
  \end{tabular}
}
\vspace{-0.2cm}
\caption{\textbf{Description of psychological tasks to predict.}}
\label{table_tasks}
\end{table}

@@

We were further interested in the utility of the
optimized low-rank projection
in one task dataset for dimensionality reduction in another task dataset.
To this end, the HCP-derived network decompositions were used as preliminary
step in the classification problem of another large sample.
The ARCHI dataset \cite{pinel07} provides activity maps from
diverse experimental tasks, including auditory and visual perception, motor action,
reading, language comprehension and mental calculation.
Analogous to HCP data, the second task dataset thus incorporated 1404
labeled, grey-matter masked, and z-scored activity maps
from 18 diverse tasks acquired in 78 participants.

\paragraph{Linear autoencoder.}
The labeled and unlabeled data were fed into a linear statistical model
composed of an autoencoder and dimensionality-reducing logistic regression.
The affine autoencoder takes the input 
$\mathbf{x}$, projects it into a coordinate system of
latent representations $\mathbf{z}$
and reconstructs it back to $\mathbf{x'}$ by
\begin{eqnarray}
    \mathbf{z} = \mathbf{W_0} \mathbf{x} + \mathbf{b_0} \hspace{2cm}
    \mathbf{x'} = \mathbf{W_1} \mathbf{z} + \mathbf{b_1},
  \label{eq:autoenc}
\end{eqnarray}

@@

%with \(f_{AE}(\mathbf x) = %
%\mathbf{W_1}(\mathbf{W_0}\mathbf{x} + \mathbf{b_0}) + \mathbf{b_1}\)
% the autoencoder.
% XXX Michael: I don't know to what extent the following makes sense or is
% XXX correct, so I am commenting it
% XXX Danilo: I found this in Bengio's papers somewhere
% This reconstruction error criterion equates with
% maximizing a lower bound on the mutual information between
% input and the learned representation.
% XXX Give a ref ?
Here we choose $\mathbf{W_0}$ and $\mathbf{W_1}$ to be tied, i.e.
$\mathbf{W_0} = \mathbf{W_1^T}$.
Consequently, the learned weights
 are forced to take a two-fold function: That of signal
\textit{analysis} and that of signal \textit{synthesis}.
The first layer \textit{analyzes}
the data to obtain the cleanest latent representation,
while the second
layer represents building blocks from which to \textit{synthesize} the data
using the latent activations.
Tying these processes together makes the analysis
layer interpretable and pulls all non-zero singular values towards 1. 
Nonlinearities were not applied to the
activations in the first layer.

\paragraph{Factored logistic regression.}
Our factored logistic regression model is best described as a
variant of a multinomial logistic regression.
Specifically, the weight
matrix is replaced by the product of two weight matrices with a common
latent dimension.
The later is typically much lower than the dimension of the data.
%
Alternatively, this model can be viewed as a single-hidden-layer
feed-forward neural network with a linear activation function for the
hidden layer and a softmax function on the output layer.
As the
dimension of the hidden layer is much lower than the input layer, this
architecture is sometimes referred to as a ``linear bottleneck'' in the
literature.
%
The probability of an input $\mathbf{x}$ to belong
to a class $i \in \{1, \dots, l\}$ is given by
\begin{equation}
  \begin{split}
    P(Y=i|\mathbf{x; V_0,V_1,c_0, c_1}) &= \softmax_i(f_\mathcal{LR}(\mathbf x)),
  \end{split}
  \label{eq:lr}
\end{equation}
where \(f_\mathcal{LR}(\mathbf x) = \mathbf{V_1 (V_0 x + c_0) + c_1}\) computes 
multinomial logits and \(\softmax_i(x) = \exp(x_i)/\sum_j\exp(x_j)\).
The matrix $\mathbf{V_0 \in \mathbb{R}^{dxn}}$
transforms the input $\mathbf{x \in \mathbb{R}^{d}}$
into $n$ latent components
and the matrix $\mathbf{V_1 \in \mathbb{R}^{nxl}}$
projects the latent components
onto hyperplanes that reflect $l$ label probabilities.
$\mathbf{c_0}$ and $\mathbf{c_1}$ are
bias vectors.
The loss function is given by
\begin{equation}
    \mathbb E\left[{\mathcal{L_{LR}}}(\mathbf{x, y})\right] \approx %
\frac{1}{N_{X_{task}}} \sum_{k=0}^{N_{X_{task}}} \log(P(Y=y^{(k)}|\mathbf{x^{(k)}; V_0, V_1, c_0, c_1}).
\label{eq:lr_loss}
\end{equation}

\paragraph{Layer combination.}
Importantly, the optimization problem of the linear autoencoder
and the factored logistic regression
are linked in two ways. First, their transformation matrices mapping from
input to the latent space are tied
\begin{eqnarray}
  \mathbf{V_0} = \mathbf{W_0}.
\end{eqnarray}
We hence search for a compression of the 79,941 voxel values into $n$ unknown
components that represent a latent code optimized for both
rest and task activity data.
Second, the objectives of the autoencoder and the factored
logistic regression are interpolated in the common loss function
% XXX Michael: In resolving conflict I opted for my change. Feel free to revert if that was intended.
\begin{equation}
{\mathcal{L}}(\theta, \lambda) = \lambda {\mathcal{L_{LR}}}
+ (1-\lambda)\frac{1}{N_{X_{rest}}} {\mathcal{L_{AE}}} + \Omega(\theta).
  \label{eq:loss_equ}
\end{equation}

@@

\paragraph{Optimization.}
The common objective was optimized
by gradient descent in the SSFLogReg parameters.
%^
The required gradients were obtained by using the chain rule to
backpropagate error derivatives.
We chose the \textit{rmsprop} solver \cite{rmsprop},
a refinement of stochastic gradient descent.
\textit{Rmsprop} dictates an adaptive learning rate
for each model parameter by
scaled gradients from a running average.
The batch size was set to $100$
(given much expected redundancy in $X_{rest}$ and $X_{task}$),
matrix parameters were initalized by Gaussian random values multiplied
by $0.004$ (i.e., gain), and
bias parameters were initalized to $0$.

The normalization factor and the update rule for $\theta$
are given by
% XXX Michael: It is crucial, but nowhere mentioned, that the optimization is
% XXX done in an alternating manner. The whole structure relies on this, by
% XXX construction it is vulnerable to optimizing too much towards labels or
% XXX rest at a time. I also identify this as one of the major weak points
% XXX of the method, so it needs to be stated in a way that is not easily
% XXX attackable.
% XXX Maybe we could also talk less about RMSprop.
%
% XXX Danilo: I think we are not alternating, Xrest and Xtask go in
% simultaneously at every bach in every epoch
\begin{eqnarray}
  \begin{split}
    \mathbf{v^{(t+1)}} &= \rho \mathbf{v^{(t)}} + (1 - \rho)\left(\nabla_{\theta} f(x^{(t)}, y^{(t)}, \theta^{(t)})\right)^2
% XXX: I have changed the first equation, please check
%v^{(t+1)} = \rho v^{(t+1)} + (1 - \rho)˜|\nabla f
\\
\theta^{(t+1)} &= \theta^{(t)} + \alpha \frac{\nabla_{\theta} f(x^{(t)}, y^{(t)}, \theta^{(t)})}{\sqrt{\mathbf{v^{(t+1)}} + \epsilon}},
  \end{split}
\end{eqnarray}
% XXX Michael: Why, in the formula, do we have two different notations for
% XXX partial derivatives? \partial and \nabla
% XXX Michael: Probably remove the explanations following and just state values
% XXX: Bertrand: what is f ?? Danilo: replaced f by more detailed notation
where $f$ is the loss function computed on a minibatch sample at timestep $t$,
$\alpha$ is the learning rate ($0.00001$),
$\epsilon$ a global damping factor ($10^{-6}$),
and
$\rho$ the decay rate ($0.9$ to deemphasize the magnitude of the gradient).
%
Note that we have also experimented with other solvers
(stochastic gradient descent, adadelta, and adagrad) but found that
\textit{rmsprop} converged faster and with
similar or higher generalization performance.
% \paragraph{Hints.}
% In fact, the constraint by a rest-data autoencoder qualifies as a
% \textit{hint}
% rather than regularization in a strict sense \cite{abu1994hints}.
% Its purpose is to introduce
% % XXX Michael: I use f earlier up, is this f here necessary?
% prior information on
% \textit{known} properties of the \textit{unknown} target function $f$.
% Rather than only relying on input-output pairs in the learning process,
% we thus narrow our hypothesis set to the biologically most plausible solutions.
% That is, we reduce the search space in a way that
% is compatible with the expected representation of BOLD activity signals.

\paragraph{Implementation.}
The analyses were performed in Python.
We used \textit{nilearn} to handle
the large quantities of neuroimaging data 
\cite{abrah14}
and
\textit{Theano} for automatic, numerically stable
differentiation of symbolic computation graphs
\cite{bastien2012theano, bergstra2010theano}.
All Python scripts that generated the results are
accessible online for reproducibility and reuse
(\url{http://github.com/anonymous/anonymous}).
% (url{http://github.com/banilo/nips2015}).

@@

SPCA separated the task-related BOLD signals into

@@

We made three noteworthy observations (Table \ref{table_two}).
%
First, the most supervised estimator ($\lambda=1$) achieved in no
instance the best accuracy, precision, recall, or f1 scores on HCP data.
Classification by SSFLogReg is therefore facilitated by
imposing structure from the unlabeled rest data.
%
Second, the higher the number of latent components $n$,
the higher the out-of-study performance with small values of $\lambda$.
This suggests that the presence of more rest-data-inspired hidden components
results in more effective feature representations in unrelated task data.
%
Third, for $n=20$ and $100$ (but not $5$) the purely rest-data-trained
decomposition matrix ($\lambda=0$) resulted in
noninferior out-of-study performance
of 77,0\% and 79,2\%,
respectively (Table \ref{table_two}).
This confirms that guiding model learning by task-unrelated structure
extracts features of general relevance
beyond the supervised problem at hand.

@@

We further evaluated whether the low-dimensional space
learned in HCP task/rest data can be re-used as a feature
extraction step for learning classification models
in an independent task dataset.
These experiments therefore probe
the generality of the learned representation
by assessing transfer learning effects.
To this end, the HCP-derived network decompositions were used as preliminary
step in the classification problem of another large sample.
The ARCHI dataset \cite{pinel07} provides activity maps from
diverse experimental tasks, including auditory and visual perception, motor action,
reading, language comprehension and mental calculation.
81 right-handed healthy participants
(3 not included in present analyses due to incomplete data)
without psychiatric or
neurological history participated in four fMRI sessions acquired under
different experimental paradigms.
The functional maps were warped into
the MNI space and resampled to isotropic 3mm resolution.
Whole-brain EPI data were acquired with the same Siemens Trio with a 32
channel head coil (TR=2400ms, TE=30ms, flip angle=60°, in-plane
FOV=$192\textrm{mm}\times192\textrm{mm}$, 40 slices, 3.0mm isotropic voxels).
Standard preprocessing was performed with Nipype \cite{gorgo11}, including
slice timing, motion correction, alignment, and spatial normalization.
Activity maps were spatially smoothed by
a Gaussian kernel of 5mm (FWHM).
Analogous to HCP data, the second task dataset incorporated 1404
labeled, grey-matter masked, and z-scored activity maps
from 18 diverse tasks acquired in 78 participants.
